Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU

### 2016-02-19 paper/paperSource.tex
automatic triggered changes

Once the data is loaded into global memory of GPU, we have two kernels for two types of queries setup for two different streams. Besides it, we allocate some memory in GPU global memory for results, too, and keep it as an object with fields to easily pull it back when it is ready, in our case all the data sizes are static before the kernels are executed, which follows nature of GPU devices. 
\subsubsection{One-body kernel} Every thread of execution in one-body query writes into shared memory, before accumulating the data into global memory result in order to avoid massive read and write blocking conflicts which will eliminate any kind of performance advantages. Once a block is finished, first thread of each block will accumulate data from shared memory into global synchronizing the results, every write operation to shared and global memories is atomic (e.g. generic CUDA function atomicAdd). We run the first stream with one-body queries and wait till it gets finished. The reason is that we expect certain behavior from shared memory from each of the streams, thus, just to make sure that they don't overlap, we run them separately. Meanwhile, while the computation is running, there is a possibility to use a third stream to load another chunk of data, too. First stream and kernel execute simple vector reduction type of computation accumulating all the needed data and keep in the GPU.
\subsubsection{Two-body kernel} Once the first kernel is finished we run the two-body kernel with its own stream. As it has already been mentioned, the biggest workload is in this kernel, because it has higher complexity. In this particular case it is $O(n^2)$. GPU streaming multiprocessors are very powerful for straightforward primitive independent calculations, and each core is much weaker than a moden CPU which has a whole bunch of L caches with some prediction behavior systems for code divergence. As you can see on Figure \ref{fg:threadload}, unbalanced thread load assumes for GPU kernel to check every time if the target atom can interact with other atom and if it can't it has to wait till other threads finish their jobs. Thus, we had to come up with a strategy of making equally balanced thread jobs and distribute them among the threads leaving atom to thread mapping as is.

Here are the steps taken through out the analysis:
\begin{itemize}
\item[(1)] Execute the data transformer
\begin{itemize}
\item[(a)] Read the MS data from trajectory files
\item[(b)] Extract the info needed for our system
\item[(c)] Save the read data to a file recognizable to the system
\end{itemize}
\item[(2)] Load the data into main memory (one frame at a time)
\begin{itemize}
\item[(a)] Load data into a double array (for one-body queries)
\item[(b)] Load data into the quad-tree structure (for two-body queries:SDH, RDF)
\end{itemize}
\item[(3)] Push the data to all queries
\item[(4)] A query, if available, acts upon the pushed data (first executing
the lower level, sub-queries)
\item[(5)] Store intermediate results (results
of sub-queries)
\item[(6)] Repeat steps 3-5 if needed.
\item[(7)] Output results
\item[(8)] Go to step 2 and load the next frame (if needed).
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
 \centerline{ \includegraphics[width=.99\columnwidth]{images/workload_set_up1.eps} }
 \caption{Workload setup}
 \label{fg:workload_set_up}
\end{figure}

\section{Experimental results}\label{sc:experiments}

The system was implemented in C++ programming language and tested on
real molecular simulation data sets. The experiments were carried
out on an Apple MacPro machine with 8GB of physical memory and two
Quad-Core Intel Xeon 3GHz processors. The MacPro was running OS X
Mavericks $10.9.3$ operating system. We have compared the results
obtained by our system to those obtained by running the analysis
through GROMACS (v. $4.5.7$). Both systems were analyzing the same
data sets.

>>>>

Once the data is loaded into global memory of GPU, we have two kernels for two types of queries setup for two different streams. Besides it, we allocate some memory in GPU global memory for results, too, and keep it as an object with fields to easily pull it back when it is ready, in our case all the data sizes are static before the kernels are executed, which follows nature of GPU devices. 
\subsubsection{One-body kernel} Every thread of execution in one-body query writes into shared memory, before accumulating the data into global memory result in order to avoid massive read and write blocking conflicts which will eliminate any kind of performance advantages. Once a block is finished, first thread of each block will accumulate data from shared memory into global synchronizing the results, every write operation to shared and global memories is atomic (e.g. generic CUDA function atomicAdd). We run the first stream with one-body queries and wait till it gets finished. The reason is that we expect certain behavior from shared memory from each of the streams, thus, just to make sure that they don't overlap, we run them separately. While the computation is running, there is a possibility to use a third stream to load another chunk of data, too. First stream and kernel execute simple vector reduction type of computation accumulating all the needed data and keep it in the GPU.
\subsubsection{Two-body kernel} Once the first kernel is finished we run the two-body kernel with its own stream. As it has already been mentioned, the biggest workload is in this kernel, because it has higher complexity. In this particular case it is $O(n^2)$. GPU streaming multiprocessors are very powerful for straightforward primitive independent calculations, and each core is much weaker than a moden CPU which has a whole bunch of L caches with some prediction behavior systems for code divergence. As you can see on Figure \ref{fg:threadload}, unbalanced thread load assumes for GPU kernel to check every time if the target atom can interact with other atom and if it has to wait till other threads finish their jobs. Thus, we had to come up with a strategy of making equally balanced thread jobs and distribute them among the threads leaving atom to thread mapping as is.

For SDH we the following strategy

\begin{itemize}
\item[1.] Find current index (for target atom)
\item[2.] Find range of atoms it has to interact with in an equally loaded fashion
\item[3.] Load the range of atoms into shared memory
\item[4.] Iterate through the range of atoms from shared memory and atomically write the distances counts into buckets pre-allocated in share memory
\item[5.] Synchronize with other threads in the block (not in the entire grid)
\item[6.] If it is the first thread in block, add all bucket counts from shared memory into global memory
\end{itemize}

\section{Experiments Results}\label{sc:experiments}

The framework has been developed in a programming language CUDA, and the sequential version for benchmark\cite{mainPaper} has been originally developed C++. Experiments were run on operating systems which didn't have any additional workload besides usual installation, and the operating systems particularly were Amazon Linux, which doesn't have a lot of modifications except for some additional features for management with Amazon EC-2 Console. We have generated mock data sets with different variations of atoms, frames and ran them with different workloads comparing sequential version of Push-based system with improved version written in CUDA.

\emph{\textbf{Hardware stack.}}
There are primarily several parts that are really important in this hardware stack. Since this is primarily big data processing, we care about processing power and memory. RAM on the node that we were running has 16GB. The server allocated from Amazon has the following CPU parameters:

\begin{center}
  \begin{tabular}{ l  l }

Architecture & x86\_64 \\  
CPU op-mode(s) & 32-bit, 64-bit \\  
Byte Order & Little Endian \\  
CPU(s) & 8 \\  
On-line CPU(s) list & 0-7 \\  
Thread(s) per core & 2 \\  
Core(s) per socket & 4 \\  
CPU family & 6 \\  
Model & 45 \\  
Model name & Intel Xeon E5-2670 2.60GHz \\  
Stepping & 7 \\  
CPU MHz & 2593.814 \\  
L1d cache & 32K \\  
L1i cache & 32K \\  
L2 cache & 256K \\  
L3 cache & 20480K \\  
NUMA node0 CPU(s) & 0-7 \\

  \end{tabular}
\end{center}

The same CPU has been used for sequential as well as for improved GPU version during experiments with benchmarks. Just to mentioned, CPU in improved version is used in-between memory loading from file and uploading it to GPU, printing the results to the user, which is not big load. Consequently, we also have GPU device with the following parameters:

\begin{center}
  \begin{tabular}{ l l }
Name & Grid 34C K520 \\
Memory & 4095Mib \\ 
Total Nvidia CUDA Cores & 1536 \\ 
Memory Size & 8 GB GDDR5 (4 GB/GPU) \\
Total GPUs & 2 GK104 GPUs \\
  \end{tabular}
\end{center}
